Stochastic blockmodels for relational events

### 2012-03-15 docs/slides/slides.tex
Update slides with posterior predictive checks material.

\begin{frame}{Stochastic blockmodels for relational event data}
Idea: 
  \begin{itemize}
  \item Assume each actor $i$ belongs to a latent class $z_i$.
  \item Model event dynamics between blocks.
  \end{itemize}

\begin{align}
\log \lambda_{ij}(t | \mathcal{A}_t,\mathbf{z}) = \boldsymbol{\beta}_{z_i,z_j} \mathbf{s}(t,i,j|\mathcal{A}_t,\mathbf{z}).
\end{align}

\uncover<2>{Notation:
\begin{itemize}
\item $\boldsymbol{\beta}_{k,l}$: parameter vector for relational events from an actor in block $k$ to an actor in block $l$
\item $\mathcal{A}_t$: history prior to time $t$
\item $\mathbf{z}$: vector of class assignments
\end{itemize}
}

\uncover<3>{
\vspace{-3cm}
Benefits:  
\begin{itemize}
\item Adjust for unobserved heterogeneity.
\item Obtain clusters of nodes who have similar interaction dynamic with other parts of the network.
\end{itemize}
}

@@

\begin{frame}{Model specification for $\mathbf{s}(t,i,j)$}
\textbf{Participation shifts} adjust for the next event
\begin{itemize}
\item AB-BA
\item AB-BY
\item AB-AY
\item AB-XA
\item AB-XB
\end{itemize}

\textbf{Degree effects} count the number of previous events
\begin{itemize}
\item sent by sender
\item received by sender
\item sent be receiver
\item received by receiver
\item previous occurrences of this dyad
\end{itemize}

@@

\begin{figure}
\centering 
%\subfigure[Dynamic network data]
%{
\begin{tikzpicture}[scale=.8,transform shape,thick]
  \node[vertex] (n1) at (0,0)  {1};
  \node[vertex] (n2) at (2,2)  {2};
  \node[vertex] (n3) at (4,0) {3};
  \node[vertex] (n4) at (2,-2) {4};
\draw[-stealth] (.4,.4) -- (1.6,1.6);
\draw[-stealth] (2.4,1.6) -- (3.6,.4);
\node at (.7,1.1) {$t_1$};
\node at (3.3,1.1) {$t_2$};
\end{tikzpicture}
% }
% \subfigure[Intensities for two dyads]
% {
\begin{tikzpicture}[scale=1,transform shape,thick]
\draw[-stealth] (0,0) -- (0,3);
\draw[-stealth] (0,0) -- (5,0);

@@

% labels
\node at (-.5,2) {$\lambda_{1,2}$};
\node at (-.5,1) {$\lambda_{3,4}$};
\node at (1,-.3) {$t_1$};
\node at (3,-.3) {$t_2$};
\end{tikzpicture}
%}
\end{figure}

\begin{itemize}
\item Restricts the class of effects one can use
\item Complexity of likelihood computation reduced ($N^2 \rightarrow N$)
\end{itemize}
\end{frame}

@@

\begin{frame}{Prediction experiment: Synthetic data}
Mean log likelihood on observed events and held out events.
\input{../../figs/synthetic/results.tex}
\end{frame}

@@

\begin{frame}{Prediction experiment: Eckmann subset}
Mean log likelihood on observed events and held out events.
\input{../../figs/eckmann-small/results.tex}
\end{frame}

@@

\uncover<3->{
Future directions
  \begin{itemize}
  \item Consider a mixed-membership approach
  \item Dirichlet process instead of fixed $K$
  \end{itemize}
}
\end{frame}

>>>>

\begin{frame}{Stochastic blockmodels for relational event data}
Idea: 
  \begin{itemize}
  \item Assume each actor $i$ belongs to a latent class $z_i$.
  \item Model event dynamics between blocks.
  \end{itemize}
\uncover<2->{
\begin{align}
\log \lambda_{ij}(t | \mathcal{A}_t,\mathbf{z}) = \boldsymbol{\beta}_{z_i,z_j} \mathbf{s}(t,i,j|\mathcal{A}_t).
\end{align}
}
\uncover<3>{Notation:
\begin{itemize}
\item $\boldsymbol{\beta}_{k,l}$: parameter vector for relational events from an actor in block $k$ to an actor in block $l$
\item $\mathcal{A}_t$: history prior to time $t$
\item $\mathbf{z}$: vector of class assignments
\end{itemize}
}

\uncover<4>{
\vspace{-3cm}
Benefits:  
\begin{itemize}
\item Adjust for unobserved heterogeneity.
\item Obtain clusters of nodes who have similar interaction dynamic with other parts of the network.
\end{itemize}
}

@@

\begin{frame}{Model specification for $\mathbf{s}(t,i,j)$}
\textbf{Participation shifts} adjust for the next event
\begin{itemize}
\item AB-BA: Reciprocity
\item AB-BY:  Turn-taking 
\item AB-AY: Turn-continuing
\item AB-XA: Turn-usurping
\item AB-XB: Turn-usurping
\end{itemize}

\textbf{Degree effects} count the number of previous events
\begin{itemize}
\item sent by sender
\item received by sender
\item sent be receiver
\item received by receiver
\item involving this dyad
\end{itemize}

@@

\uncover<2->{
\begin{figure}
\centering 
%\subfigure[Dynamic network data]
%{
\begin{tikzpicture}[scale=.8,transform shape,thick]
  \node[vertex] (n1) at (0,0)  {1};
  \node[vertex] (n2) at (2,2)  {2};
  \node[vertex] (n3) at (4,0) {3};
  \node[vertex] (n4) at (2,-2) {4};
\draw[-stealth] (.4,.4) -- (1.6,1.6);
\draw[-stealth] (2.4,1.6) -- (3.6,.4);
\node at (.7,1.1) {$t_1$};
\node at (3.3,1.1) {$t_2$};
\end{tikzpicture}
% }
% \subfigure[Intensities for two dyads]
% {
\begin{tikzpicture}[scale=1,transform shape,thick]
\draw[-stealth] (0,0) -- (0,3);
\draw[-stealth] (0,0) -- (5,0);

@@

% labels
\node at (-.5,2) {$\lambda_{1,2}$};
\node at (-.5,1) {$\lambda_{3,4}$};
\node at (1,-.3) {$t_1$};
\node at (3,-.3) {$t_2$};
\end{tikzpicture}
%}
\end{figure}
}
\uncover<3->{
\begin{itemize}
\item Restricts the class of effects one can use
\item Complexity of likelihood computation reduced ($N^2 \rightarrow N$)
\end{itemize}
}
\end{frame}

@@

\begin{frame}{Posterior predictive checks}
  \begin{itemize}
  \item Investigate whether observed data is reasonable under model.
  \item Interested in a particular statistic of a sequence $T(Y)$ \\
    (e.g. total time, degree distribution, etc.)
  \item Simulate $Y^{(i)} \sim \mbox{REM}(\beta^{(i)},\mathcal{A})$
  \item Compare $T(Y)$ to distribution of $T(Y^{(i)})$
  \end{itemize}
\end{frame}

\begin{frame}{Posterior predictive checks}
\includegraphics[scale=.4]{../../figs/synthetic/postpred-time}
\includegraphics[scale=.4]{../../figs/synthetic/postpred-pshift}
\end{frame}

\begin{frame}{Prediction experiment: Synthetic data}
Mean log likelihood for observed events and held out events.
\input{../../figs/synthetic/results.tex}
\end{frame}

@@

\begin{frame}{Prediction experiment: Eckmann subset}
Mean log likelihood for observed events and held out events.
\input{../../figs/eckmann-small/results.tex}
\end{frame}

\begin{frame}{Posterior predictive checks for $K=2$ model}
\includegraphics[scale=.4]{../../figs/eckmann-small/postpred-time}
\includegraphics[scale=.4]{../../figs/eckmann-small/postpred-pshift}
\end{frame}

@@

\uncover<3->{
Future directions
  \begin{itemize}
  \item Consider a mixed-membership approach
  \item Intelligent way to share parameters as $K$ grows
  \end{itemize}
}
\end{frame}
